---
title: "Plotting results returned by functions in sfhotspot"
vignette: >
  %\VignetteIndexEntry{Plotting sfhotspot results}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
fig-align: center
fig-asp: 0.7
fig-width: 7
---

The `hotspot_*()` functions return simple-feature (SF) objects with specialised classes. `autoplot()` recognises those classes and chooses a variable, colour scale, and legend suited to the result. Some choices also depend on the columns or attributes present in the result.


| Result | Default variable shown on map |
|---|---|
| `hotspot_count()` (`hspt_n`)                | `sum` when weighted counts are present; otherwise `n` |
| `hotspot_kde()` (`hspt_k`)                  | `kde` |
| `hotspot_dual_kde()` (`hspt_dk`)            | Scale selected from the `method` attribute |
| `hotspot_change()` (`hspt_d`)               | Diverging scale of `change`, centred on zero |
| `hotspot_classify()` (`hspt_c`)             | Categorical hotspot and coldspot colours |
| `hotspot_gistar()` (`hspt_g`) with `kde`    | Density in statistically significant cells |
| `hotspot_gistar()` (`hspt_g`) without `kde` | Diverging scale of `gistar`, centred on zero |
| `hotspot_dbscan()` (`hspt_db`)                | `n` |

The maps below show examples of the default maps produced for each `hotspot_*()` function, based on the `memphis_robberies`, `memphis_precincts` and `memphis_population` datasets included in the package. To get started, we create a grid covering the city of Memphis and transform the dataset to use an appropriate projected coordinate reference system (CRS), since that makes specifying cell size, etc., easier.

```{r}
#| label: setup

library(sf)
library(sfhotspot)
library(ggplot2)

# Transform objects to use a local co-ordinate reference system
memphis_robberies_proj <- st_transform_auto(memphis_robberies)  
memphis_precincts_proj <- st_transform_auto(memphis_precincts, quiet = TRUE)
memphis_population_proj <- st_transform_auto(memphis_population, quiet = TRUE)

# Set up consistent grids of cells
memphis_grid <- hotspot_grid(memphis_precincts_proj, cell_size = 500)
```


::: {.callout-note title="Grid size"}
The 500-metre grid cells used in these examples may be too coarse for many analytical purposes, but are used here because of constraints on rendering time for R package vignettes. In practice, the cell size should be chosen to suit the spatial scale of the data and the research question.
:::


In each case the `autoplot()` method loads a map with no base map by default. For each of the above functions a corresponding `hotspot_map()` method exists that by-default also includes a base map with suitable attribution, etc. `hotspot_map()` can also be used for mapping ordinary `sf` objects.


## `hotspot_count()`

For an unweighted point count, `autoplot()` maps the `n` column.

```{r}
#| label: plot-unweighted-count
#| fig-alt: >
#|   Grid map of Memphis robberies. Darker blue cells contain more robberies.

memphis_robberies_count <- hotspot_count(
  memphis_robberies_proj, 
  grid = memphis_grid
)

autoplot(memphis_robberies_count)
```

If `hotspot_count()` is given weights, its result also contains a `sum` column. The plotting methods detect that column and map the weighted count instead.

`hotspot_layer()` applies the same variable mapping as `autoplot()`, but returns
a layer so that it can be combined with other ggplot2 layers. It is a
student-friendly wrapper around `ggplot2::autolayer()`. For example, a count
layer can be drawn with precinct boundaries.

```{r}
#| label: plot-count-hotspot-layer
#| fig-alt: >
#|   Map of Memphis robbery counts drawn with outlines of police precincts.
#|   Areas of greater robbery counts are darker blue.

ggplot() +
  hotspot_layer(memphis_robberies_count) +
  geom_sf(data = memphis_precincts_proj, fill = NA, colour = "grey70") +
  scale_fill_distiller(palette = "Blues", direction = 1) +
  labs(fill = "Robbery\ncount") +
  theme_void()
```

`hotspot_layer()` is particularly useful because it allows control of where the
layer derived from the result produced by a `hotspot_*()` function appears in a
stack of other spatial layers, so that (for example) it can be placed on top of
a base map.


## `hotspot_kde()`

When provided with an object produced by `hotspot_kde()`, `autoplot()` maps the `kde` column. The default colour scale is sequential, with darker blue indicating greater density. Note that since `hotspot_kde()` produces density estimates for the convex hull of the data, the resulting map may extend beyond the precinct boundaries. `hotspot_clip()` can be used to restrict the KDE to the specific area covered by the point dataset to avoid displaying spurious density estimates for areas not covered by the data.

```{r}
#| label: plot-kde
#| cache: true
#| fig-alt: >
#|   Map of the density of robberies in Memphis. Darker blue areas have a
#|   higher density of robberies.

memphis_robberies_proj |> 
  hotspot_kde(bandwidth_adjust = 0.25, grid = memphis_grid) |> 
  hotspot_clip(memphis_precincts_proj) |> 
  autoplot()
```

## `hotspot_dual_kde()`

`hotspot_dual_kde()` can compare the density of two spatial datasets in four
ways. The method is recorded in the `method` attribute of the result, which
`autoplot()` and `hotspot_layer()` use to select the appropriate colour scale. The
following examples use the same robbery and population data, bandwidth and grid
so that the effect of changing only `method` is visible.

The default, `method = "ratio"`, divides the density of the first dataset by
the density of the second. This is analogous to mapping a rate. The map uses a
sequential scale applied directly to the ratio values, without treating one as
a change point or transforming the scale.

```{r}
#| label: plot-dual-kde-ratio
#| cache: true
#| fig-alt: >
#|   Map of the density of robberies relative to the population density in 
#|   Memphis. Darker blue areas have a higher density of robberies relative to 
#|   the population.

hotspot_dual_kde(
  memphis_robberies_proj, 
  memphis_population_proj, 
  method = "ratio",
  bandwidth_adjust = 0.25,
  grid = memphis_grid
) |>
  autoplot()
```

`method = "log"` calculates the natural logarithm of the density ratio. Zero
then represents equal densities, negative values indicate a lower robbery
density relative to population density, and positive values indicate a higher
relative robbery density. Although zero denotes equality, it is not treated as
a colour-scale boundary: a continuous sequential scale represents progression
from lower to higher logged ratios.

```{r}
#| label: plot-dual-kde-log
#| cache: true
#| fig-alt: >
#|   Sequential map of the logged ratio of robbery density to population
#|   density in Memphis. Lighter and darker areas have lower and higher logged
#|   density ratios.

hotspot_dual_kde(
  memphis_robberies_proj,
  memphis_population_proj,
  method = "log",
  bandwidth_adjust = 0.25,
  grid = memphis_grid
) |>
  autoplot()
```

`method = "diff"` subtracts the density of the second dataset from that of the
first. This is the only method for which `autoplot()` uses a diverging scale:
negative and positive differences are shown on opposite sides of zero.

```{r}
#| label: plot-dual-kde-difference
#| cache: true
#| fig-alt: >
#|   Diverging map of robbery density minus population density in Memphis.
#|   Negative and positive density differences appear on opposite sides of the
#|   colour scale.

hotspot_dual_kde(
  memphis_robberies_proj,
  memphis_population_proj,
  method = "diff",
  bandwidth_adjust = 0.25,
  grid = memphis_grid
) |>
  autoplot()
```

Finally, `method = "sum"` adds the two density surfaces. Since sums are
non-negative, they are shown with a sequential scale: darker cells have a
higher combined density.

```{r}
#| label: plot-dual-kde-sum
#| cache: true
#| fig-alt: >
#|   Sequential map of the combined robbery and population density in Memphis.
#|   Darker blue cells have a higher combined density.

hotspot_dual_kde(
  memphis_robberies_proj,
  memphis_population_proj,
  method = "sum",
  bandwidth_adjust = 0.25,
  grid = memphis_grid
) |>
  autoplot()
```

Each legend uses `low` at its lower end and `high` at its upper end. These
labels describe position within the values produced by the selected method;
they do not imply that the four methods produce directly comparable numeric
values.


## `hotspot_change()`

`hotspot_change()` returns the differences between the number of points in each area in two time periods. Its scale is symmetric around zero so increases and decreases of the same size receive equal visual emphasis.

```{r}
#| label: plot-change
#| fig-alt: >
#|   Diverging grid map of changes in robbery counts between the first and
#|   second halves of 2019. Decreases and increases use opposite colours.

memphis_robberies_proj |>
  hotspot_change(
    boundary = as.POSIXct("2019-07-01", tz = "UTC"),
    grid = memphis_grid
  ) |>
  autoplot()
```


## `hotspot_classify()`

`hotspot_classify()` categorises grid cells based on their hotspot and coldspot status over time. The default palette uses warm hues for hotspot categories, cool hues for coldspot categories, purple for mixed results, and grey for no pattern.

```{r}
#| label: plot-classification
#| fig-alt: >
#|   Categorical map classifying Memphis grid cells as persistent, emerging,
#|   intermittent or former hotspots and coldspots, mixed areas, or no pattern.

memphis_robberies_proj |>
  hotspot_classify(period = "1 month", grid = memphis_grid) |>
  autoplot()
```


## `hotspot_gistar()`

By default, `hotspot_gistar()` returns density estimates and p-values for each cell alongside the Getis-Ord Gi* statistic. The plotting methods use the p-values to select which cells are visible, then map the density of points in each visible cell. The `critical_p` and `sign` arguments to `autoplot()`/`hotspot_layer()` can control this behaviour. Cells with `pvalue >= critical_p` are transparent. The default is `critical_p = 0.05`. The `sign` argument can be used to restrict the visible cells to those with significant positive or negative Gi*/Gi values. When both signs are shown, a diverging scale distinguishes cold spots from hot spots; cold-spot densities are negated only for plotting. When only hot spots or only cold spots are shown, a medium-to-dark blue scale ensures that every significant cell remains visible.

```{r}
#| label: plot-gistar-both
#| fig-alt: >
#|   Density map showing only areas with significantly more or fewer robberies
#|   than expected by chance. Other grid cells are transparent.

memphis_robberies_proj |>
  hotspot_gistar(grid = memphis_grid, bandwidth_adjust = 0.25) |>
  autoplot()
```

If `kde = FALSE`, the result has no `kde` column. The plotting methods instead map the `gistar` column using a diverging scale centred on zero. In this case, the `critical_p` and `sign` arguments to `autoplot()`/`hotspot_layer()` are ignored.

```{r}
#| label: plot-gistar-without-kde
#| fig-alt: >
#|   Diverging grid map of Getis-Ord Gi-star statistics for Memphis robberies,
#|   centred on zero. Positive and negative values use opposite colours.

memphis_robberies_proj |>
  hotspot_gistar(grid = memphis_grid, bandwidth_adjust = 0.25, kde = FALSE) |>
  autoplot()
```


## `hotspot_dbscan()`

`hotspot_dbscan()` returns polygons covering clusters of points identified by the DBSCAN algorithm. By default, the `fill` aesthetic is mapped to the `n` column, which counts the number of points in each cluster. This can be changed using the `col_fill` argument to `autoplot()` or `hotspot_layer()`. Polygons can also be labelled using the `col_label` argument to `autoplot()` or `hotspot_layer()`. The default colour scale is sequential, with darker blue indicating clusters with more points.


```{r}
#| label: plot-dbscan
#| fig-alt: >
#|   Map of clusters of robberies in Memphis identified by the DBSCAN algorithm.

memphis_robberies_proj |>
  hotspot_dbscan(density_adjust = 3) |>
  autoplot()
```


## Isobands

The numeric grid outputs from `hotspot_count()`, `hotspot_kde()`,
`hotspot_dual_kde()`, `hotspot_change()` and `hotspot_gistar()` can be
generalised into isobands using `hotspot_isoband()`. The categorical output
from `hotspot_classify()` cannot be converted because isobands require numeric
values on a regular square grid.

Isobands retain information about the original `hotspot_*()` result and can
themselves be plotted using `autoplot()` or `hotspot_layer()`. The default palette,
legend and treatment of meaningful values such as zero therefore depend on the
function that produced the original grid. For example, KDE isobands use a
sequential scale:

```{r}
#| label: plot-kde-isobands
#| fig-alt: >
#|   Isoband map of robbery density in Memphis. Darker blue bands represent
#|   areas with higher estimated robbery density.

memphis_robberies_proj |>
  hotspot_kde(bandwidth_adjust = 0.25, grid = memphis_grid) |>
  hotspot_isoband() |>
  autoplot()
```


## Customising plots

`autoplot()` returns an ordinary ggplot object, so scales, labels, themes, and
other layers can be added or replaced with ggplot2. `hotspot_layer()` is useful
when the sfhotspot result is one component of a larger map. See the individual
method help pages for the arguments accepted by each method.

An optional map-tile background can provide geographic context. This requires
the suggested ggspatial package and an internet connection unless the tiles are
already cached. When a base map is requested, the hotspot layer is drawn with
an alpha of 0.75 so that the background remains visible.

```{r}
#| eval: false

autoplot(
  memphis_robberies_count,
  basemap_type = "osm",
  basemap_zoom = 12
)
```

The default remains `basemap_type = "none"`, so ordinary plots never download
tiles.
Known map types receive an attribution caption automatically. For a custom or
new map type, supply the provider's required statement explicitly using
`basemap_attribution`. Use `caption` to add contextual information such as the
map author, date or notes. This text appears above any explanatory caption and
base-map attribution, which are retained automatically.
